---
title: "Untitled"
author: "Michael Rose"
date: "April 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(keras)
library(reticulate)
use_python('/home/michael/anaconda3/bin/python3')
```

Training a NN revolves around the following objects: 
  1. Layers, which are combined into a network (or model)
  2. Input data and corresponding targets
  3. loss function which defines the feedback signal used for learning
  4. optimizer which determines how learning proceeds
  
# Layers - The Building Blocks of Deep Learning

The typical keras workflow looks something like: 

  1. Define your training data (input tensors and target tensors)
  2. Define a network of layers (or model) that maps your inputs to your targets
  3. Configure the learning process by choosing a loss function, an optimizer, and some metrics to monitor
  4. Iterate on your training data by calling the fit() method of your model

```{r}

# step 2
# linear stacks of layers
model <- keras_model_sequential() %>% 
  layer_dense(units = 32, input_shape = c(784)) %>% 
  layer_dense(units = 10, activation = "softmax")

# same model with functional api
input_tensor <- layer_input(shape = c(784))
output_tensor <- input_tensor %>% 
  layer_dense(units = 32, activation = "relu") %>% 
  layer_dense(units = 10, activation = "softmax")

model <- keras_model(inputs = input_tensor, outputs = output_tensor)

# with the functional api we are manipulating data tensors that the model processes and applying layers to this tensor as if they were functions

# step 3
# example using a single loss function

model %>% compile(
  optimizer = optimizer_rmsprop(lr = 0.0001),
  loss = "mse",
  metrics = c("accuracy")
)

# step 4
model %>% fit(input_tensor, target_tensor, batch_size = 128, epochs = 10)

```

# IMDB

```{python}
from keras.datasets import imdb
imdb = imdb.load_data()
# imdb_word_index = imdb.get_word_index()
```

```{r}
# %<-% is the multiassignment operator from the zeallot package. It allows us to unpack the list into a set of distinct variables
imdb <- dataset_imdb(num_words = 10000)
c(c(train_data, train_labels), c(test_data, test_labels)) %<-% imdb

# alternatively, the code above pans out to 

imdb <- dataset_imdb(num_words = 10000)
train_data <- imdb$train$x
train_labels <- imdb$train$y
test_data <- imdb$test$x
test_labels <- imdb$test$y

```

```{r}
# show encoding indices
str(train_data[[1]])

# decode words back to english
word_index <-dataset_imdb_word_index() # named list mapping words to integer index
reverse_word_index <- names(word_index) # reverses it, mapping integer indices to words
names(reverse_word_index) <- word_index
decoded_review <- sapply(train_data[[1]], function(index) {
  word <- if (index >= 3) reverse_word_index[[as.character(index - 3)]]
  if (!is.null(word)) word else "?"
})

```

You can't just feed lists of integers into a neural network - we have to turn them into tensors first. Either: 
  1. Pad your lists so that they all have the same length, turn them into an integer tensor of shape (samples, word_indices) 
  2. One-shot encode our lists into vectors of 0s and 1s - e.g. [3,5] is a tensor with all 0s except for 3 and 5 which are ones

```{r}
# encoding sequences into a binary matrix (choice 2)
vectorize_sequences <- function(sequences, dimension = 10000) {
  results <- matrix(0, nrow = length(sequences), ncol = dimension) # creates an all zero matrix of shape (length(sequences), dimension)
  for (i in 1:length(sequences))
    results[i, sequences[[i]]] <- 1 # set specific indices of results[i] to 1s
  results
}

x_train <- vectorize_sequences(train_data)
x_test <- vectorize_sequences(test_data)

str(x_train[1,]) # take a look at the data

# convert labels from integer to numeric
y_train <- as.numeric(train_labels)
y_test <- as.numeric(test_labels)
```

```{r}
# create model. returns a probability by ending the network with the sigmoid function

model <- keras_model_sequential() %>% 
  layer_dense(units = 16, activation = "relu", input_shape = c(10000)) %>% 
  layer_dense(units = 16, activation = "relu") %>% 
  layer_dense(units = 1, activation = "sigmoid")

# choose a loss function and optimizer

model %>% compile(
  optimizer = "rmsprop",
  loss = "binary_crossentropy",
  metrics = c("accuracy")
)

# sometimes we may want to tweak the compilation a bit: 

# configure optimizer parameters: 

model %>% compile(
  optimizer = optimizer_rmsprop(lr=0.001),
  loss = "binary_crossentropy",
  metrics = c("accuracy")
)

# using custom losses and metrics
model %>% compile(
  optimizer = optimizer_rmsprop(lr = 0.001),
  loss = loss_binary_crossentropy, # function argument
  metrics = metric_binary_accuracy # function argument
)



 ```

